Biden to win the popular vote in the 2020 American presidential elections in a closely fought contest

### 2020-11-03 outputs/paper/paper.Rmd
Updated abstract

---
title: "Biden to win the popular vote in the 2020 American presidential elections"
author: "Anne Collins, Jennifer Do, Andrea Therese Javellana, Wijdan Tariq"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
abstract: \noindent "Why do/should we care about the problem and the results. What problem are we trying to solve. How did we go about solving or making progress on the problem? What is our answer? What are the implications of our answer?."
indent: true
---

@@

@@

While the ACS data measures hundreds of variables, we extracted 10 demographic variables we thought could predict the popular vote in the 2020 American presidential election. They were: age, sex, household income, education attainment, employment status, state, region, birthplace, race, and whether a respondent had Hispanic origins. These variables were also chosen because of their ability to coincide with that of the survey variables. For example, household income was chosen rather than individual income since the UCLA survey did not ask about personal income. We further manipulated the data by cleaning responses to match the UCLA options. For income, we constructed bins of income intervals that match the survey’s since the ACS had exact income values. We selected respondents between 18 and 93 years of age. We made birthplace a binary variable to be either born in the ‘USA’ or ‘another country’. We constructed a new race variable that incorporates Hispanic origins; this meant that if a respondent had answered they had Hispanic origins, it would override and replace the answer of their identifying race. We also kept Chinese identifying respondents separate from other Asians and did these things because Chinese and Hispanic respondents have shown to have strong voting trends with contemporary topics like America’s border policies and COVID-19 (Krogstad and Lopez, 2020). The figure below depicts the variables of the ACS data after cleaning.

>>>>

---
title: "Biden to win the popular vote in the 2020 American presidential elections in a closely fought contest"
author: "Anne Collins, Jennifer Do, Andrea Therese Javellana, Wijdan Tariq"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
abstract: \noindent "The 2020 American presidential elections are the most discussed elections in the world due to their potential and profound impact on global affairs. In this paper, we aim to predict whether Biden or Trump will win the popular vote and by how much of a margin. We ran a logistic regression model using 5,127 observations from election survey data provided by Democracy Fund + UCLA Netscape. To provide more robust predictions, we utilize the multilevel regression with poststratification method using over 2 million observations from the American Community Survey (ACS) data. While the survey data suggests that Biden will get 52% of the popular vote compared to Trump's 48%, our post-stratification results suggest that the election will be even closer, with 51% predicted for Biden and 49% for Trump. We conclude the paper with a discussion of the benefits of using post-stratification on survey data to correct for potential sample biases and to get more reliable regression estimates."
indent: true
---

@@

The survey dataset we used was retrieved from the Nationscape Data Set. This dataset was created in partnership between the Democracy Fund Voter Study Group and UCLA Political Scientists Chris Tausanovitch and Lynn Vavreck (Tausanovitch and Vavreck, 2020). Nationscape had conducted surveys to 500,000 Americans from July 2019 to December 2020 in lead up to the 2020 campaign and election. Each week, the survey team had interviewed roughly 6,520 people. To provide access to audiences, Lucid, a market research platform, had provided samples and an online exchange for survey respondents for Nationscape. The samples taken from this exchange included demographic quotas including age, gender, ethnicity, region, income and education. All respondents had conducted the survey online, along with an attention check prior to responding to the survey. The survey team interviewed people across the U.S, and had accounted for respondents in nearly all counties, congressional districts, and mid-sized U.S cities. 
To ensure accurate representation of the American population, the survey data was weighted and generated using a raking technique, with weights generated per week’s surveys. The weights were derived from the 2017 American Community Survey of the U.S Census Bureau’s adult population and its respective demographics (such as gender, region, race, household income, education, age, 2016 presidential vote, etc.). As well, representativeness was followed according to the Pew Research Center’s evaluations of non-probability samples. The Nationscape results were compared to results from the 2018 and 2016 Pew Reports. It was determined that the Nationscape estimates were close to samples by the Pew Research Center.

The data set we used was obtained from the Phase 2 of Nationscape’s data, and accounted for the Nationscape Wave 50, taken from June 25th to July 1st, 2020. As the Nationscape data set had hundreds of variables, we chose to extract 11 variables focusing on demographics. These variables included: age, gender, employment, race, Hispanic ethnicity, household income, state, census region, education, nativity (where the respondent was born), and their choice in vote between Donald Trump and Joe Biden in the upcoming 2020 election. Variables such as race and Hispanic ethnicity were combined into a new race variable, to account for Hispanic ethnicity as a race. As well, responses to voter choice which did not include Trump or Biden as answers were deleted, to align with our primary goal of predicting the winner of the 2020 election between these two primary candidates. We had also created bins for variables race, education, and employment to match the cleaned ACS data set for post stratification (further discussed later on). From these variables, we chose to further analyze 5 of these variables (age, gender, employment, race, and voter choice) to coincide with the variables analyzed in the ACS data set.

@@

While the ACS data measures hundreds of variables, we extracted 10 demographic variables we thought could predict the popular vote in the 2020 American presidential election. They were: age, sex, household income, education attainment, employment status, state, region, birthplace, race, and whether a respondent had Hispanic origins. These variables were also chosen because of their ability to coincide with that of the survey variables. For example, household income was chosen rather than individual income since the UCLA survey did not ask about personal income. We further manipulated the data by cleaning responses to match the UCLA options. For income, we constructed bins of income intervals that match the survey’s since the ACS had exact income values. We selected respondents between 18 and 93 years of age. We made birthplace a binary variable to be either born in the ‘USA’ or ‘another country’. We constructed a new race variable that incorporates Hispanic origins; this meant that if a respondent had answered they had Hispanic origins, it would override and replace the answer of their identifying race. We also kept Chinese identifying respondents separate from other Asians and did these things because Chinese and Hispanic respondents have shown to have strong voting trends with contemporary topics like America’s border policies and COVID-19 (Krogstad and Lopez, 2020).

The figures below compare the variables of the ACS data after cleaning with the survey data. From the plots generated from the survey data, it was observed that the majority of respondents were aged between 30 to 44 years, and were split evenly between male and female respondents. The majority of respondents were observed to be employed at 57% of the respondent population, followed by respondents not actively in the work force at 34% (Figure 3). The majority of respondents were White (69%), followed by Native American (14%), and Black Americans (10%), and other races ranged between 1-3% of the respondent population (Figure 4). Based on the survey, a greater distribution of respondents were seen to be more likely to vote for Joe Biden (Democratic candidate) in the upcoming 2020 election compared to Donald Trump (Republican candidate), at 51% and 48% of respondents answering respectively (Figure 5). This may be due to respondent bias, as the Nationscape dataset itself was in partnership with the Democracy Fund, it is expected that a greater number of respondents who support the democratic party would respond.
